Cross-validating mixed-effects models

### 2024-06-30 scratch/notes_georges/vignettes_drafts/cv-mixed_modified-edited.Rmd
cv-mixed with John's 2024-06-29 edits

Consider a researcher studying the effect of the dosage of a drug on a measure of the severity of symptoms for some hypothetical disease. The researcher has longitudinal data on 20 patients, each of whom was observed on five occasions in which patients received different dosages of the drug.  The data are observational with dosages prescribed by the patients' physicians so that patients who were more severely affected by the disease received higher dosages of the drug.

A figure below shows the four simulated contrived data sets illustrating possible results for data obtained in such a scenario. The relative configuration of dosages, 'x', and symptoms, 'y', are identical within patients in each of the four datasets. Within patients, higher dosages are generally associated with a reduction in symptoms.

Between patients, however, higher dosages are associated with higher levels of symptoms. A plausible mechanism would be a reversal of causality: within patients, higher dosages may alleviate symptoms, but, between patients, higher morbidity may cause the prescription of higher dosages.

The four data sets differ in the between-patient variance of patient centroids from a regression line. They exhibit a progression from low to high variance around the regression line.  We start by generating a data set to serve as a common template for the four sample data sets.  We then apply different multipliers to the between-patient variation using parameters consistent with the description above o highlight the issues that arise in cross-validating mixed-effects models:[^1]

[^1]: We invite the interested reader to experiment with varying the parameters of our example.

```{r include=FALSE, echo=FALSE}
library(glmmTMB) # necessary for some reason to knit vignette in RStudio, harmless otherwise
```
```{r}

@@

SD_between <- c(0,5,6,8)                  # SD between patients
SD_within <- rep(2.5, length(SD_between)) # SD within patients

Nv <- length(SD_within)       # number of variance profiles
SD_ratio <- paste0(SD_between,' / ',SD_within)
SD_ratio <- factor(SD_ratio, levels = SD_ratio)

@@

Data_template <- expand.grid(patient = 1:Nb, obs = 1:Nw) |>
  within({
    xw <- seq(-2,2, length.out = Nw)[obs]
    x <- patient + xw
    xm  <- ave(x, patient)   # within-patient mean

@@

    re_std <- scale(resid(lm(rnorm(Nb*Nw)~ x)))
    re_between <- ave(re_std, patient)
    re_within <- re_std - re_between
    re_between <- scale(re_between)/sqrt(Nw)
    re_within <- scale(re_within)
  })

@@

Data <- within(
  Data,
  {
    SD_within_ <- SD_within[SD_ratio_i]
    SD_between_ <- SD_between[SD_ratio_i]
    SD_ratio <- SD_ratio[SD_ratio_i]
    y <- 10 +
      Bb * xm +                  # contextual effect
      Bw * (x - xm) +            # within-patient effect
      SD_within_ * re_within +   # within patient random effect
      SD_between_ * re_between   # adjustment to between patient random effect
  }
)
```
Here is a scatterplot of the data sets showing estimated 50% concentration ellipses for each cluster:[^2]

[^2]: We find it convenient to use the **lattice** [@Sarkar:2008] and **latticeExtra** [@SarkarAndrews:2022] packages for this and other graphs in this section.

```{r plot1}
#| out.width = "100%",
#| fig.height = 4.5,
#| fig.cap = "Data sets showing identical within-patient configurations with increasing between-patient variance. The labels above each panel show the between-patient SD/within-patient SD."
library("lattice")
library("latticeExtra")
plot <- xyplot(y ~ x | SD_ratio, data=Data, group=patient,
           layout = c(Nv,1),
           par.settings=list(superpose.symbol=list(pch=1, cex=0.7))) +
      layer(panel.ellipse(..., center.pch = 16, center.cex = 1.5,  level = 0.5))
plot # display graph
```
The population from which these datasets are generated has a between-patient effect of dosage of 1 and a within-patient effect of -0.5. The researcher may attempt to obtain an estimate of the within-patient effect of dosage through the use of mixed models with a random intercept. We will illustrate how this approach results in estimates that are highly sensitive to the relative variance at the two levels of the mixed model by considering four models, all of which include a random intercept.  The first three models have sequentially nested fixed effects models: 1) intercept only, `~ 1`, 2) intercept and effect of 'x', `~ 1 + x` and 3) intercept, effect of 'x' and a contextual variable, 'xm', consisting of the within-patient mean of 'x', `~ 1 + x + xm`. The fourth model uses an intercept and the 'centered-within-group' variable 'x - xm', `~ 1 + I(x = xm)`. We thus fit four models to four datasets for a total of sixteen models.

```{r}
model.formulas <- c(
  intercept =  y ~ 1 + (1 | patient),
  x         =  y ~ 1 + x + (1 | patient),
  'x + xm'  =  y ~ 1 + x + xm + (1 | patient),
  '(x-xm)'  =  y ~ 1 + I(x - xm) + (1 | patient)
)

fits <- lapply(split(Data, ~ SD_ratio),
               function(d) {
                 lapply(model.formulas, function(form) {
                   glmmTMB(form, data = d)
                 })
               })
```
We proceed to obtain predictions from each model based on fixed effects alone, as would be used for cross-validation based on clusters (i.e., patients), and for fixed and random effects---so-called best linear unbiased predictions or BLUPs---as would be used for cross-validation based on cases (i.e., occasions within patients):
```{r}
pred.BLUPs <- lapply(fits, lapply, predict)                # predicted fixed and random effects
pred.fixed <- lapply(fits, lapply, predict, re.form = ~0)  # predicted fixed effects
```
We then prepare the data for plotting:
```{r}
Dataf <- lapply(split(Data, ~ SD_ratio),
                    function(d) {
                      lapply(names(model.formulas), function(form) cbind(d, formula = form))
                    }) |> 
             lapply(function(dlist) do.call(rbind, dlist) ) |> 
             do.call(rbind, args = _)

@@

The fixed-effects predictions from these models are shown in the following graph:
```{r plot-fits-mod1}
#| out.width = "100%",
#| fig.height = 6,
#| fig.cap = "Fixed-effect predictions using each model applied to data sets with varying between- and within-patient variance ratios."
xyplot(pred.fixed  ~ x |SD_ratio * formula, Dataf, type = 'l', groups = patient,
       ylab = 'fixed-effect predictions') |> 
  useOuterStrips() |> print()
```
The BLUPs from these models are shown in the following graph:
```{r plot-fits-mod2}
#| out.width = "100%",
#| fig.height = 6,
#| fig.cap = "Fixed- and random-effect predictions (BLUPs) using each model applied to data sets with varying between- and within-patient variance ratios."
xyplot(pred.BLUPs  ~ x |SD_ratio * formula, Dataf, type = 'l', groups = patient,
       ylab = 'fixed- and random-effect predictions (BLUPS)') |> 
  useOuterStrips() |> print()
```
Data sets with relatively low between-patient variance result in strong shrinkage of fixed-effects predictions and also of BLUPs towards the between-patient relationship between 'y' and 'x' in the model with an intercept and 'x' as fixed-effect predictors, `~ 1 + x`. The inclusion of a contextual variable in the model corrects this problem.

@@

Comparing the '5 / 2.5' dataset with the '8 / 2.5' dataset, we will note that, with the former dataset, case-based cross-validation prefers the `~ 1 + x` model to the `~ 1` model but cluster-based cross-validation prefers the latter to the former. The situation is entirely reversed with the '8 / 2.5' dataset.

The third model, `~ 1 + x + xm`, includes a contextual effect of $x$---that is, the cluster mean `xm`---along with $x$ and the intercept in the fixed-effect part of the model, and a random intercept. This model is equivalent to fitting `y ~ I(x - xm) + xm + (1 | patient)`, which is the model that generated the data.  The fit of the mixed model `~ 1 + x + xm` is consequently similar to that of a fixed-effects only model with 'x' and a categorical predictor for individual patients (i.e., `y ~ factor(patient) + x`, treating patients as a factor, and not shown here).

@@

cvs_clusters <-
  lapply(1:Nv,
         function(i){
           cv(model_lists[[i]],  k = 10, data = split(Data, ~SD_ratio)[[i]], clusterVariables = 'patient')
         })
```
For a given dataset, we can plot the results for a list of models using the plot method for 'cvList'. For example
```{r}
#| out.width = "60%",
#| fig.height = 3,
#| fig.cap = "10-fold cluster-based cross-validation comparing random intercept mixed models with varying fixed effects."
plot(cvs_clusters[[2]])
```
We assemble the results for all datasets to show them in a common figure.
```{r}
names(cvs_clusters) <- names(cvs_cases) <- SD_ratio 
dsummary <- expand.grid(SD_ratio_i = names(cvs_cases), model = names(cvs_cases[[1]]))

@@

dsummary$clusters <-
  sapply(1:nrow(dsummary), function(i){
    with(dsummary[i,], cvs_clusters[[SD_ratio_i]][[model]][['CV crit']])
  })
```
```{r}
#| out.width = "100%",
#| fig.height = 4,
#| fig.cap = "10-fold cluster- and case-based cross-validation comparing mixed models with random intercepts and various fixed effects."
xyplot(cases + clusters ~ model|SD_ratio_i, dsummary,
       auto.key = list(space = 'top', reverse.rows = T, columns = 2), type = 'b',
       ylab = 'CV criterion (MSE)',
       layout= c(Nv,1),
       par.settings =
         list(superpose.line=list(lty = c(2, 3), lwd = 3),
              superpose.symbol=list(pch = 15:16, cex = 1.5)),
       scales = list(y = list(log = TRUE), x = list(alternating = F, rot = 60))) |> print()
```
In summary, when between-cluster variance is relatively large, model `~ 1 + x`, with $x$ alone and without the contextual mean of $x$, is assessed as fitting very poorly by cluster-based CV, but relatively much better by case-based CV. In all our examples, model `~ 1 + x + xm`, which includes both $x$ and its contextual mean, produces better results using both cluster-based and case-based CV.  These conclusions are consistent with our observations based on graphing predictions from the various models, and they illustrate the desirability of assessing mixed-effect models at different hierarchical levels.

>>>>

Consider a researcher studying the effect of the dosage of a drug on the severity of symptoms for a hypothetical disease. The researcher has longitudinal data on 20 patients, each of whom was observed on five occasions in which patients received different dosages of the drug.  The data are observational, with dosages prescribed by the patients' physicians, so that patients who were more severely affected by the disease received higher dosages of the drug.

Our four contrived data sets (see below) illustrate possible results for data obtained in such a scenario. The relative configuration of dosages `x` and symptoms `y` are identical within patients in each of the four datasets. Within patients, higher dosages are generally associated with a reduction in symptoms.

Between patients, however, higher dosages are associated with higher levels of symptoms. A plausible mechanism is a reversal of causality: Within patients, higher dosages alleviate symptoms, but between patients higher morbidity causes the prescription of higher dosages.

The four data sets differ in the between-patient variance of patient centroids from a common between-patients regression line. The data sets exhibit a progression from low to high variance around the common regression line.

We start by generating a data set to serve as a common template for the four sample data sets.  We then apply different multipliers to the between-patient variation using parameters consistent with the description above to highlight the issues that arise in cross-validating mixed-effects models:[^data-template]

[^data-template]: We invite the interested reader to experiment with varying the parameters of our example.

```{r include=FALSE, echo=FALSE}
library("glmmTMB") # necessary for some reason to knit vignette in RStudio, harmless otherwise
```
```{r}

@@

SD_between <- c(0, 5, 6, 8)               # SD between patients
SD_within <- rep(2.5, length(SD_between)) # SD within patients

Nv <- length(SD_within)       # number of variance profiles
SD_ratio <- paste0('SD ratio = ', SD_between,' / ',SD_within)
SD_ratio <- factor(SD_ratio, levels = SD_ratio)

@@

Data_template <- expand.grid(patient = 1:Nb, obs = 1:Nw) |>
  within({
    xw <- seq(-2, 2, length.out = Nw)[obs]
    x <- patient + xw
    xm  <- ave(x, patient)   # within-patient mean

@@

    re_std <- scale(resid(lm(rnorm(Nb*Nw) ~ x)))
    re_between <- ave(re_std, patient)
    re_within <- re_std - re_between
    re_between <- scale(re_between)/sqrt(Nw)
    re_within <- scale(re_within)
  })

@@

Data <- within(
  Data,
  {
    SD_within_ <- SD_within[SD_ratio_i]
    SD_between_ <- SD_between[SD_ratio_i]
    SD_ratio <- SD_ratio[SD_ratio_i]
    y <- 10 +
      Bb * xm +                  # contextual effect
      Bw * (x - xm) +            # within-patient effect
      SD_within_ * re_within +   # within patient random effect
      SD_between_ * re_between   # adjustment to between patient random effect
  }
)
```
Here is a scatterplot of the data sets showing estimated 50% concentration ellipses for each cluster:[^lattice]

[^lattice]: We find it convenient to use the **lattice** [@Sarkar:2008] and **latticeExtra** [@SarkarAndrews:2022] packages for this and other graphs in this section.

```{r plot1}
#| out.width = "100%",
#| fig.height = 4.5,
#| fig.cap = "Data sets showing identical within-patient configurations with increasing between-patient variance. The labels above each panel show the between-patient SD/within-patient SD. The population between-patient regression line, $E(y) = 10 + x$, is shown in each panel."
library("lattice")
library("latticeExtra")
plot <- xyplot(y ~ x | SD_ratio, data=Data, group=patient,
           layout = c(Nv, 1),
           par.settings=list(superpose.symbol=list(pch=1, cex=0.7))) +
      layer(panel.ellipse(..., center.pch = 16, center.cex = 1.5,  
                          level = 0.5),
            panel.abline(a=10, b=1))
plot # display graph
```
The population from which these datasets is generated has a between-patient effect of dosage of 1 and a within-patient effect of $-0.5$. The researcher may attempt to obtain an estimate of the within-patient effect of dosage through the use of mixed models with a random intercept. We will illustrate how this approach results in estimates that are highly sensitive to the relative variance at the two levels of the mixed model by considering four models, all of which include a random intercept.  The first three models have sequentially nested fixed effects: (1)`~ 1`, intercept only; (2)`~ 1 + x`, intercept and effect of `x`; and (3) `~ 1 + x + xm`, intercept, effect of `x`, and a contextual variable, `xm`, consisting of the within-patient mean of `x`. The fourth model, `~ 1 + I(x - xm)`, uses an intercept and the centered-within-group variable `x - xm`. We thus fit four models to four datasets for a total of 16 models.

```{r}
model.formulas <- c(
  ' ~ 1'             =  y ~ 1 + (1 | patient),
  '~ 1 + x'          =  y ~ 1 + x + (1 | patient),
  '~ 1 + x + xm'     =  y ~ 1 + x + xm + (1 | patient),
  '~ 1 + I(x - xm)'  =  y ~ 1 + I(x - xm) + (1 | patient)
)

fits <- lapply(split(Data, ~ SD_ratio),
               function(d) {
                 lapply(model.formulas, function(form) {
                   glmmTMB(form, data = d)
                 })
               })
```
We proceed to obtain predictions from each model based on fixed effects alone, as would be used for cross-validation based on clusters (i.e., patients), and for fixed and random effects---so-called best linear unbiased predictions or "BLUPs"---as would be used for cross-validation based on cases (i.e., occasions within patients):
```{r}
# predicted fixed and random effects:
pred.BLUPs <- lapply(fits, lapply, predict)
# predicted fixed effects:
pred.fixed <- lapply(fits, lapply, predict, re.form = ~0)  
```
We then prepare the data for plotting:
```{r}
Dataf <- lapply(split(Data, ~ SD_ratio),
                    function(d) {
                      lapply(names(model.formulas), 
                             function(form) cbind(d, formula = form))
                    }) |> 
             lapply(function(dlist) do.call(rbind, dlist) ) |> 
             do.call(rbind, args = _)

@@

The fixed-effects predictions from these models are shown in the following graph:
```{r plot-fits-mod1}
#| out.width = "100%",
#| fig.height = 6,
#| fig.cap = "Fixed-effect predictions using each model applied to data sets with varying between- and within-patient variance ratios."
xyplot(pred.fixed  ~ x |SD_ratio * formula, Dataf, type = 'l', groups = patient,
       ylab = 'fixed-effect predictions') |> 
  useOuterStrips() |> print()
```
The BLUPs from these models are shown in the following graph:
```{r plot-fits-mod2}
#| out.width = "100%",
#| fig.height = 6,
#| fig.cap = "Fixed- and random-effect predictions (BLUPs) using each model applied to data sets with varying between- and within-patient variance ratios."
xyplot(pred.BLUPs  ~ x |SD_ratio * formula, 
       Dataf, type = 'l', groups = patient,
       ylab = 'fixed- and random-effect predictions (BLUPS)') |> 
  useOuterStrips() |> print()
```
Data sets with relatively low between-patient variance result in strong shrinkage of fixed-effects predictions and also of BLUPs towards the between-patient relationship between `y` and `x` in the model with an intercept and `x` as fixed-effect predictors, `~ 1 + x`. The inclusion of a contextual variable in the model corrects this problem.

@@

We will discover that case-based cross-validation prefers the `~ 1 + x` model to the `~ 1` model for the '5 / 2.5' dataset, but cluster-based cross-validation prefers the latter model to the former. The situation is entirely reversed with the '8 / 2.5' dataset.

The third model, `~ 1 + x + xm`, includes a contextual effect of `x`---that is, the cluster mean `xm`---along with `x` and the intercept in the fixed-effect part of the model, along with a random intercept. This model is equivalent to fitting `y ~ I(x - xm) + xm + (1 | patient)`, which is the model that generated the data.  The fit of the mixed model `~ 1 + x + xm` is consequently similar to that of a fixed-effects only model with `x` and a categorical predictor for individual patients (i.e., `y ~ factor(patient) + x`, treating patients as a factor, and not shown here).

@@

cvs_clusters <-
  lapply(1:Nv,
         function(i){
           cv(model_lists[[i]],  k = 10, 
              data = split(Data, ~SD_ratio)[[i]], 
              clusterVariables = 'patient')
         })
```
For a given dataset, we can plot the results for a list of models using the plot method for `"cvList"` objects. For example
```{r}
#| out.width = "60%",
#| fig.height = 3,
#| fig.cap = "10-fold cluster-based cross-validation comparing random intercept mixed models with varying fixed effects."
plot(cvs_clusters[[2]])
```
We assemble the results for all datasets to show them in a common figure.
```{r}
names(cvs_clusters) <- names(cvs_cases) <- SD_ratio 
dsummary <- expand.grid(SD_ratio_i = names(cvs_cases), model = names(cvs_cases[[1]]))

@@

dsummary$clusters <-
  sapply(1:nrow(dsummary), function(i){
    with(dsummary[i,], cvs_clusters[[SD_ratio_i]][[model]][['CV crit']])
  })
```
```{r}
#| out.width = "100%",
#| fig.height = 4,
#| fig.cap = "10-fold cluster- and case-based cross-validation comparing mixed models with random intercepts and various fixed effects."
xyplot(cases + clusters ~ model|SD_ratio_i, dsummary,
       auto.key = list(space = 'top', reverse.rows = T, columns = 2), type = 'b',
       ylab = 'CV criterion (MSE)',
       layout= c(Nv,1),
       par.settings =
         list(superpose.line=list(lty = c(2, 3), lwd = 3),
              superpose.symbol=list(pch = 15:16, cex = 1.5)),
       scales = list(y = list(log = TRUE), x = list(alternating = F, rot = 60))) |> print()
```
In summary, when between-cluster variance is relatively large, the model `~ 1 + x`, with `x` alone and without the contextual mean of `x`, is assessed as fitting very poorly by cluster-based CV, but relatively much better by case-based CV. In all our examples, the model `~ 1 + x + xm`, which includes both `x` and its contextual mean, produces better results using both cluster-based and case-based CV.  These conclusions are consistent with our observations based on graphing predictions from the various models, and they illustrate the desirability of assessing mixed-effect models at different hierarchical levels.
